---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

# Classification

## Introduction to Classification

In general classification is concerend with the following problem: our population consists of several distinct groups. We have a set of data points from each group with associated measurements. We want to derive a procedure that tells us which group a new observation might belong to.

#### **Example**: Fisher's Iris Data

One of the most famous data sets in Statistics was first studied by Fisher, his [iris](https://en.wikipedia.org/wiki/Iris_flower_data_set) data. For each of three types of iris flowers (Iris setosa, Iris virginica and Iris versicolor) we have four measurements: the lengths and the widths of the Petal and the Sepal. The goal is to determine from these measurements the type of flower.

```{r}
head(iris)
```

```{r}
pushViewport(viewport(layout = grid.layout(3, 3)))
print(ggplot(data=iris, 
          aes(Sepal.Length, Sepal.Width, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=1, layout.pos.col=1))
print(ggplot(data=iris, 
         aes(Sepal.Length, Petal.Length, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=1, layout.pos.col=2))
print(ggplot(data=iris, 
             aes(Sepal.Length, Petal.Width, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=1, layout.pos.col=3))
print(ggplot(data=iris, 
             aes(Sepal.Width, Petal.Length, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=2, layout.pos.col=2))
print(ggplot(data=iris, 
             aes(Sepal.Width, Petal.Width, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=2, layout.pos.col=3))
print(ggplot(data=iris, 
             aes(Petal.Length, Petal.Width, color=Species)) +
      geom_point() + theme(legend.position="none"),
  vp=viewport(layout.pos.row=3, layout.pos.col=3))
```


#### **Example**: 

Consider the following artificial examples:

-  two types, fairly simple separation

```{r}
library(mvtnorm)
ex1 <- function(mu=2, n=50) {
  x1 <- rmvnorm(n, mean=c(0,0), sigma=diag(2)) 
  x2 <- rmvnorm(n, mean=c(mu,mu), sigma=diag(2)) 
  data.frame(x=c(x1[, 1], x2[, 1]), 
             y=c(x1[, 2], x2[, 2]), 
             group=rep(c("A", "B"), each=n))
}
```

```{r}
ggplot(data=ex1(n=150), aes(x, y,color=group)) +
  geom_point() +
  theme(legend.position="none")
```

-  two types, more complicated separation

```{r}
ex2 <- function(mu=2, n=50) {
  x <- cbind(runif(10000), runif(10000, -1, 1))
  x <- x[x[, 1]^2 + x[, 2]^2<1, ]
  x <- x[1:n, ]
  y <- cbind(runif(10000, 0, 2), runif(10000, -2, 2))
  y <- y[y[, 1]^2 + y[, 2]^2>0.9, ]
  y <- y[1:n, ]
  data.frame(x=c(x[, 1], y[, 1]), 
             y=c(x[, 2], y[, 2]), 
             group=rep(c("A", "B"), each=n))
}
```

```{r}
ggplot(data=ex2(n=150), aes(x, y,color=group)) +
  geom_point() +
  theme(legend.position="none")
```

-  three types

```{r}
ex3 <- function(mu=2, n=33) {
  x1 <- rmvnorm(n, mean=c(0, 0), sigma=diag(2)) 
  x2 <- rmvnorm(n, mean=c(mu, mu), sigma=diag(2)) 
  x3 <- rmvnorm(n, mean=2*c(mu, mu), sigma=diag(2))
  data.frame(x=c(x1[, 1], x2[, 1], x3[, 1]), 
             y=c(x1[, 2], x2[, 2], x3[, 2]), 
             group=rep(c("A", "B", "C"), each=n))
}
```

```{r}
ggplot(data=ex3(n=150), aes(x, y,color=group)) +
  geom_point() +
  theme(legend.position="none")
```

`r hl()$hr()`

In one sense this is not a new problem, it is simply a regression problem where the response variable is discrete.

For this we could code a response variable y as 0 if "green" and 1 if "red" if there are two groups (models 1 and 2) or with 0, 1 and 2 if there are three groups (model 3). Then we run the linear regression of y on x1 and x2.

Finally we assign a point (x1, x2) to "green" if its predicted response is <0.5, and to "red" otherwise for models 1 and 2, and depending on whether its predicted response is <2/3 or >4/3 for model 3.

Of course in the case of two groups we could also use *logistic regression*, but we won't pursue this idea here.

To see what this looks like we find an even spaced grid and predict the color for each point. Then we overlay that grid onto the graph. This is done in

```{r}
make.grid <- function(df) {
  x <- seq(min(df$x), max(df$x), length=100)
  y <- seq(min(df$y), max(df$y), length=100)
  expand.grid(x=x, y=y) 
}
do.graph <- function(df, df1) {
  print(ggplot(data=df, aes(x, y, color=group)) +
   geom_point(size=2) +
   theme(legend.position="none") +
   geom_point(data=df1, 
             aes(x,y, color=group, alpha=0.1),
             inherits.aes=FALSE)) 
} 
```

Here our three examples:

```{r}
df <- ex1()
df$Code <- ifelse(df$group=="A", 0, 1)
fit <- lm(Code~x+y, data=df)
df1 <- make.grid(df)
df1$group <- ifelse(predict(fit, df1)<0.5, "A", "B")
do.graph(df, df1)
```


```{r}
df <- ex2()
df$Code <- ifelse(df$group=="A", 0, 1)
fit <- lm(Code~x+y, data=df)
df1 <- make.grid(df)
df1$group <- ifelse(predict(fit, df1)<0.5, "A", "B")
do.graph(df, df1) 
```

```{r}
df <- ex3()
df$Code <- ifelse(df$group=="A", 0, 1)
df$Code[df$group=="C"] <- 2
fit <- lm(Code~x+y, data=df)
df1 <- make.grid(df)
tmp <- predict(fit, df1)
df1$group <- ifelse(tmp<2/3, "A", "B")
df1$group[tmp>4/3] <-"C"
do.graph(df, df1)
```

this seems to work ok for examples 1 and 3, not so much for 2.

we will use these examples quite a bit, so lets write a routine that generates data from any of them:

```{r}
gen.ex <- function(which, n=50) {
  if(which==1) 
      df <- ex1(n=n)
  if(which==2) 
      df <- ex2(n=n)  
  if(which==3) 
      df <- ex3(n=n)  
  df$Code <- ifelse(df$group=="A", 0, 1)
  if(which==3) 
    df$Code[df$group=="C"] <- 2
  df
}
```


Let's have a closer look at example 1:

```{r}
df <- gen.ex(1)
fit <- lm(Code~x+y, data=df)
coef(fit)
```

we assign the group depending if the fitted value is < or > than 0.5. What do we get if it is equal to 0.5?

$$
\begin{aligned}
& 0.5   = \beta_0 +\beta_1 x+ \beta_2 y\\
& y   = (0.5 - \beta_0 -\beta_1 x)/\beta_2 \\
& y = \frac{0.5-\beta_1}{\beta_2}-\frac{\beta_1}{\beta_2}x
\end{aligned}
$$
Let's add that line to the graph:

```{r}
ggplot(data=df, aes(x, y, color=group)) +
  geom_point(size=2) +
  theme(legend.position="none") +
  geom_abline(intercept = (0.5-coef(fit)[2])/coef(fit)[3],
              slope=-coef(fit)[2]/coef(fit)[3])
```

and this is called the *decision boundary*.

It is easy to see that in example 3 it works like this:

```{r}
df <- gen.ex(3)
fit <- lm(Code~x+y, data=df)
ggplot(data=df, aes(x, y, color=group)) +
  geom_point(size=2) +
  theme(legend.position="none") +
  geom_abline(intercept = (2/3-coef(fit)[2])/coef(fit)[3],
              slope=-coef(fit)[2]/coef(fit)[3]) +
  geom_abline(intercept = (4/3-coef(fit)[2])/coef(fit)[3],
              slope=-coef(fit)[2]/coef(fit)[3])
```

### Misclassification Rate

One thing that sets a classification problem apart from regression is that here we have an obvious way to judge how good a method is, namely the **miss-classification rate**: What percentage of the observations are given the wrong label?

Let's see:

```{r}
msr <- function(x, y) {
  z <- table(x, y)
  round((sum(z)-sum(diag(z)))/sum(z)*100, 1)
}
```


```{r}
df <- gen.ex(1, n=1000)
fit <- lm(Code~x+y, data=df)
pred <- ifelse(predict(fit)<0.5, "A", "B")
table(df$group, pred)
msr(df$group, pred)
```


```{r}
df <- gen.ex(2, n=1000)
fit <- lm(Code~x+y, data=df)
pred <- ifelse(predict(fit)<0.5, "A", "B")
msr(df$group, pred)
```


```{r}
df <- gen.ex(3, n=1000)
fit <- lm(Code~x+y, data=df)
tmp <- predict(fit)
pred <- ifelse(tmp<2/3, "A", "B")
pred[tmp>4/3] <- "C"
msr(df$group, pred)
```

### Overfitting and Cross-validation

Of course these misclassification rates are to optimistic: we calculated it on the same data set that we fit on. We should always *train* and *test* on different data sets, maybe using cross-validation:

```{r}
df <- gen.ex(1, n=1000)
out <- rep(0, 10)
for(i in 1:10) {
  I <- sample(1:2000, size=400)
  fit <- lm(Code~x+y, data=df[-I, ])
  pred <- ifelse(predict(fit, df[I, 1:2])<0.5, "A", "B")
  out[i] <- msr(df$group[I], pred) 
}
mean(out)
```

Here we split the data into $80\%$ for training and $20\%$ for evaluation. Is this a good split? Actually, nobody knows!

`r hl()$hr()`

Our method works quite well for examples 1 and 3, but not so much for example 2.

```{r}
df <- gen.ex(2)
df$Code <- ifelse(df$group=="A", 0, 1)
fit <- lm(Code~x+y, data=df)
df1 <- make.grid(df)
df1$group <- ifelse(predict(fit, df1)<0.5, "A", "B")
do.graph(df, df1)
```

shows us why: here a linear decision boundary clearly won't work. So how about a quadratic one?

```{r}
df$x2 <- df$x^2
df$y2 <- df$y^2
df$xy <- df$x*df$y
fit <- lm(Code~x+y+x2+y2+xy, data=df)
df1 <- make.grid(df)
df1$x2 <- df1$x^2
df1$y2 <- df1$y^2
df1$xy <- df1$x*df1$y
df1$group <- ifelse(predict(fit, df1)<0.5, "A", "B")
do.graph(df, df1)
```

and that looks much better!

Here is the mcr based on cross-validation:

```{r}
df <- df[, c(4, 1:2, 5:7)]
out <- rep(0, 10)
for(i in 1:10) {
  I <- sample(1:2000, size=400)
  fit <- lm(Code~x+y+x2+y2+xy, data=df[-I, ])
  pred <- ifelse(predict(fit, df[I, -1])<0.5, "A", "B")
  out[i] <- msr(df$Code[I], pred) 
}
mean(out)
```

