---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  pdf_document:
    fig_caption: no
  html_document: default
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

# General Statistics

## Descriptive Statistics

In general in Statistics we distinguish between quantitative (= numerical) and categorical (= qualitative) data. The main difference is that for quantitative data doing arithmetic makes sense, for example calculating the mean. 

Note that just because a data set has digits, it is not necessarily quantitative. For example, digits are often used as labels.

`r hl()$hr()`

Consider the **upr admissions** data. Here are some simple things to do when looking at this kind of data:

### Tables

```{r}
Gender <- table(upr$Gender)
names(Gender) <- c("Female", "Male")
Percentage <- round(Gender/sum(Gender)*100, 1)
cbind(Gender, Percentage)
```

```{r echo=FALSE}
rm(Gender)
rm(Percentage)
```

### Contingency Tables

```{r}
tbl <- table(upr$Gender, upr$Class.Facultad)
tbl
```

In a contingency table percentages can be calculated in three ways:


```{r}
# overall total
ot <- sum(tbl)
ot
# row total
rt <- apply(tbl, 1, sum)
rt
# column total
ct <- apply(tbl, 2, sum)
ct
```

-  by grand total

```{r}
tmp <- cbind(tbl, Total=rt)
tmp <- rbind(tmp, Total=c(ct, sum(ct)))
round(tmp/ot*100, 1)
```

-  by row total

```{r}
round(tmp/c(rt, ot)*100, 1)
```

-  by column total

```{r}
t(round(t(tmp)/c(ct, ot)*100, 1))
```

### Bar Charts

```{r}
ggplot(upr, aes(Class.Facultad)) + 
  geom_bar(alpha=0.75, fill="lightblue") +
  xlab("")
ggplot(upr, aes(Class.Facultad, fill=Gender)) + 
  geom_bar(position="dodge", alpha=0.75) 
```

as with the tables, graphs can be done based on percentages:

-  grand total

```{r}
ggplot(upr, aes(Class.Facultad, fill=Gender)) + 
  geom_bar(aes(y = (..count..)/sum(..count..)),
           position="dodge", 
           alpha=0.75) 
```

-  row total

this one we have to work a bit:

```{r}
tmp1 <- c(tmp[1, 1:5], tmp[2, 1:5])/c(rt, rt)*100
df <- data.frame(Percentage = tmp1,
                 Gender=rep(c("Female", "Male"), 5),
                 Class=names(tmp1))
ggplot(df, aes(x = Class, 
               y = Percentage,
               fill = Gender)) + 
    geom_bar(position = "dodge",
             stat = "identity") 
```

Notice the use of *stat="identity"* if the data is already in the form of a table.

### Numerical Summaries

```{r error=TRUE}
round(mean(upr$Freshmen.GPA), 3)
```

we get an error because there are missing values, so 

```{r}
round(mean(upr$Freshmen.GPA, na.rm=TRUE), 3)
round(median(upr$Freshmen.GPA, na.rm=TRUE), 3)
round(sd(upr$Freshmen.GPA, na.rm=TRUE), 3)
round(quantile(upr$Freshmen.GPA, 
               probs = c(0.1, 0.25, 0.75, 0.9),
               na.rm=TRUE), 3)
```

### Histogram and Boxplot

```{r}
bw <- diff(range(upr$Freshmen.GPA, na.rm = TRUE))/50 # use about 50 bins
ggplot(upr, aes(Freshmen.GPA)) +
  geom_histogram(color = "black", 
                 fill = "white", 
                 binwidth = bw) + 
  labs(x = "Freshmen GPA", y = "Counts")
```

```{r}
ggplot(upr, aes(x="", y=Freshmen.GPA)) + 
  geom_boxplot() + 
  xlab("")
ggplot(upr, aes(factor(Year), Freshmen.GPA)) + 
  geom_boxplot() 
```

### Two Quantitative Variables

```{r}
round(cor(upr$Year, upr$Freshmen.GPA, 
    use="complete.obs"), 3)
ggplot(upr, aes(Year, Freshmen.GPA)) + 
  geom_jitter(shape=".", width=0.1, height = 0) +
  scale_x_continuous(breaks = 2003:2013) +
  labs(x="Year", y="GPA after Freshmen Year")
```

### Normal Probability Plot

An important graph is the *normal probability plot*, which plots the sample quantiles vs the population quantiles of a normal distribution:

```{r}
x <- rnorm(20)
df <- data.frame(x=x)
y1 <- quantile(x, c(0.25, 0.75))
x1 <- qnorm(c(0.25, 0.75))
slope <- diff(y1)/diff(x1)
int <- y1[1L] - slope * x1[1L]
ggplot(df, aes(sample=x)) +
  stat_qq() +
  geom_abline(slope = slope, intercept = int, color="blue")
```

Notice that adding the customary line through the quartiles takes a bit of work.

Here an example where the normal assumption fails:

```{r}
x <- rexp(20)
df <- data.frame(x=x)
y1 <- quantile(x, c(0.25, 0.75))
x1 <- qnorm(c(0.25, 0.75))
slope <- diff(y1)/diff(x1)
int <- y1[1L] - slope * x1[1L]
ggplot(df, aes(sample=x)) +
  stat_qq() +
  geom_abline(slope = slope, intercept = int, color="blue")
```

