---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  pdf_document:
    fig_caption: no
  html_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## Nonlinear Parametric Models

Sometimes the model we wish to fit is known, up to parameters. Generally that is the case if there is a scientific theory that predicts the shape of the relationship. For example, radioactive decay is known to be exponential: $y=\alpha e^{-\beta t}$

#### **Example**: Growth of Lobsters

Data from an experiment to raise Florida lobster in a controlled environment. The data shows the overall length and the age of a certain species of lobster.

```{r}
ggplot(data=lobster, aes(Time, Length)) +
  geom_point()
```

Now biology suggests that the relationship should be of the form

$$
y=\frac{\beta_2}{1+(\beta_2 - \beta_0)/\beta_0 \exp (\beta_1 t)} + \epsilon
$$
where

-  $\beta_0$ is the expected value of y at time t=0  
-  $\beta_1$ is a measure of the growth rate  
-  $\beta_2$ is the expected limit of y as $t\rightarrow \infty$

This is often called the **logistic** or **autocatalytic** model

How do we fit such a model, that is find "optimal" values of $\beta_0$, $\beta_1$ and $\beta_2$? Sometimes it is possible use transformations to "linearize" the model, for example we have of course $\log (y)= \log(\alpha)-\beta t$ for the radioactive decay model. This is not possible, though, for the logistic model, so we have to find a different solution.

Previously we have always used the method of least squares to estimate the parameters in our models, that is we minimized the "figure of merit"

$$
\text{RSS} = \sum (y_i - \beta_0 - \beta_1 x_i)^2
$$
the natural extension of this is to use

$$
\text{RSS} = \sum (y_i - f(x_i; \boldsymbol{\beta}  ))^2
$$

now for a linear model minimizing this expression could be done with *lm*. This however is still a minimiziation problem, and we can do it  with 

```{r}
fit <- nls(Length ~ beta[3]/(1 + ((beta[3] -
          beta[1])/beta[1]) *  exp(beta[2] * Time)), 
          start = list(beta = c(10, -0.1, 500)), 
          data = lobster)
summary(fit)
```

```{r}
x <- seq(10, 160, 1)
df <- data.frame(x=x, 
                 y = predict(fit, 
                  newdata = data.frame(Time = x)))
```

```{r}
 ggplot(data=lobster, aes(Time, Length)) +
  geom_point() +
  geom_line(data=df, aes(x, y), color="blue")
```
                
#### **Example**: Prime Number Theorem

That there were infinitely many prime numbers was first proven by the Greek mathematician Euclid at around 300BC. A serious study of how fast they grow was first begun by [Adrienne-Marie Legendre](https://www.britannica.com/biography/Adrien-Marie-Legendre). He studied the function N(k), which gives the number of primes less or equal to k. We can do the same. The primes up to 1,000,000 are available at

```{r}
primes <- scan("C://Users//Wolfgang//dropbox//teaching//Computing-with-R//primes.txt")
primes <- as.integer(primes)
```

A detailed study of these primes led Legendre in 1798 to propose the function

$$
N(k)=k/(\log k - \alpha)
$$
Here is what that looks like for several values of $\alpha$:

```{r}
N <- function(k, alpha) {
  k/(log(k)-alpha)
}
k <- seq(1000, 1e6, length=250)
exact.counts <- k
for(i in 1:250) 
  exact.counts[i] <- sum(primes<k[i])
df <- data.frame(N=c(k, k, k, k), 
      Counts=c(exact.counts, N(k, 0), N(k, 1), N(k, 2)),
      Method=rep(c("Counts", "a=0", "a=1", "a=2"),
                 each=250))
ggplot(df, aes(N, Counts, color=Method)) +
  geom_line()
```

and so it seems a value of $\alpha=1$ is good.

Legendre however was not satisfied with that, he wanted to find the optimal answer. So he found the least squares solution!

```{r}
fit <- nls(exact.counts ~ k/(log(k) - alpha), 
          start = list(alpha = 0))
coef(fit)
```

and so he claimed that 

$$
N(k)=k/(\log k - 1.08)
$$

Around the same time German mathematician [Carl Friedrich Gauss](https://www.britannica.com/biography/Carl-Friedrich-Gauss) also looked at this problem, and he made a different conjecture. He said

$$
N(k)=k/\log k 
$$
That was a rather strange guess, because it looks like this:

```{r}
N <- function(k, alpha) {
  k/(log(k)-alpha)
}
k <- seq(1000, 1e6, length=250)
exact.counts <- k
for(i in 1:250) 
  exact.counts[i] <- sum(primes<k[i])
df <- data.frame(N=c(k, k), 
      Counts=c(exact.counts, N(k, 0)),
      Method=rep(c("Counts", "Gauss"),   each=250))
ggplot(df, aes(N, Counts, color=Method)) +
  geom_line()
```

and it surely looks like the two curves are growing further apart. However, almost 100 years later in 1896 the French mathematicians Jacques-Salomon Hadamard and Charles de la Valée Poussin independently showed that Gauss was right!

From our modern point of view we might say Legendre was guilty  of **over-fitting**!

