---
title:  R Basics I
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%")
library(ggplot2)
library(grid)
```

`r hl()$basefontsize()`

To start run

```{r, eval=FALSE}
ls() 
```

This shows you a "listing"" of the files (data, routines etc.) in the current roject. (Likely there is nothing there right now)

Everything in R is either a data set or a function. It is a function if it is supposed to do something (maybe calculate something, show you something like a graph or something else etc. ). If it is a function is ALWAYS NEEDS (). Sometimes the is something in between the prentices, like in the hplot() above. Sometimes there isn't like in the ls(). But the () has to be there anyway.

If you have worked for a while you might have things you need to save, do that by clicking on 

File > Save 

RStudio has a nice recall feature, using the up and down arrow keys. Also, clicking on the History tab shows you the recently run commands. Finally, typing the first three letters of a command in the console and then typing CTRL-^  shows you a list of when you ran commands like this the last times.

R is case-sensitive, so a and A are two different things.

Often during a session you create objects that you need only for a short time. When you no longer need them use  **rm** to get rid of them:
  
```{r}
x <- 10
x^2
rm(x) 
```

the **<-** is the *assignment* character in R, it assigns what is on the right to the symbol on the left. (Think of an arrow to the left)

##  Data Entry 

For a few numbers the easiest thing is to just type them in:
```{r}
x <-  c(10, 2, 6, 9)
x
```
c() is a function that takes the objects inside the () and  **c**ombines them into one single object (a vector). 

## Data Types in R

the most basic type of data in R is a **vector**, simply a list of values. 

Say we want the numbers 1.5, 3.6, 5.1 and 4.0 in an R vector called x, then we can type 

```{r}
x <- c(1.5, 3.6, 5.1, 4.0)
x
```

Often the numbers have a structure one can make use of:

``` {r}
1:10 
10:1
1:20*2
c(1:10, 1:10*2)
```

Sometimes you need parentheses:

```{r}
n <- 10
1:n-1
1:(n-1)
```

The *rep* ("repeat") command is very useful:

```{r}
rep(1, 10)
rep(1:3, 10)
rep(1:3, each=3)
rep(c("A", "B", "C"), c(4,7,3))
```

what does this do?
```{r, eval=FALSE}
rep(1:10, 1:10)
```

## Commands for Vectors

To find out how many elements a vector has use the *length* command: 

```{r}
x <- c(1.4, 5.1, 2.0, 6.8, 3.5, 2.1, 5.6, 3.3, 6.9, 1.1)
length(x)
```

The elements of a vector are accessed with the bracket [ ] notation:

```{r}
x[3]
x[1:3]
x[c(1, 3, 8)]
x[-3]
x[-c(1, 2, 5)]
```

Instead of numbers a vector can also consist of characters (letters, numbers, symbols etc.) These are identified by quotes:

````{r} 
c("A", "B", 7, "%")
```

A vector is either numeric or character, but never both (see how the 7 was changed to "7").

You can turn one into the other (if possible) as follows:

```{r}
x <- 1:10
x
as.character(x)
```

```{r}
x <- c("1", "5", "10", "-3")
x
as.numeric(x)
```

A third type of data is logical, with values either TRUE or FALSE. 

```{r}
x <- 1:10
x
x > 4
```

these are often used as conditions:

```{r}
x[x>4]
```

This, as we will see shortly, is EXTREMELY useful!

## Data Frames

data frames are the basic format for data in R. They are essentially vectors put together as columns. 

The main thing you need to know about working with data frames are the following commands:

### Case Study: UPR Admissions

consider the  **upr** data set . This is the application data for all the students who applied and were accepted to UPR-Mayaguez between 2003 and 2013. 

```{r}
dim(upr)
```

tells us that there were 23666 applications and that for each student there are 16 pieces of information.

```{r}
colnames(upr)
```

shows us the variables 

``` {r}
head(upr, 3)
```

shows us the first three cases.

Let's say we want to find the number of males and females. We can use the table command for that:

```{r,echo=FALSE}
rm(Gender)
```


``` {r, eval=FALSE}
table(Gender)
```

```{r, echo=FALSE}
cat("Error: object 'Gender' not found\n")
```


What happened? Right now R does not know what Gender is because it is "hidden" inside the upr data set. Think of **upr** as a box that is currently closed, so R can't look inside and see the column names. We need to open the box first:

```{r}
attach(upr)
table(Gender)
```

there is also a detach command to undo an attach, but this is not usually needed because the attach goes away when you close R.  

**Note**: you need to attach a data frame only once in each session working with R. 

**Note**: Say you are working first with a data set "students 2016" which has a column called Gender, and you attached it. Later (but in the same R session) you start working with a data set "students 2017" which also has a column called Gender, and you are attaching this one as well. If you use Gender now it will be from "students 2017".

## Subsetting of Data Frames

Consider the following data frame (not a real data set):

```{r, echo=FALSE}
set.seed(111)
Age <- sample(18:25, size=10, replace=T)
GPA <- round(runif(10, 2, 4), 1)
Gender <- sample(c("Male", "Female"), size=10, replace=T)
students <- data.frame(Age, GPA, Gender)
```

```{r}
students
```

Here each single piece of data is identified by its row number and its column number. So for example in row 2, column 2 we have "3.2", in row 6, column 3 we have "Male". 

As with the vectors before we can use the  [ ] notation to access pieces of a data frame, but now we need to give it both the row and the column number, separated by a ,:

```{r}
students[6, 3]
```

As before we can pick more than one piece:

```{r}
students[1:5, 3]
students[1:5, 1:2]
students[-c(1:5), 3]
students[1, ]
students[, 2]
students[, -3]
```

## Vector Arithmetic 

R allows us to apply any mathematical functions to a whole vector:

```{r}
x <- 1:10
2*x
```

```{r}
x^2
```

```{r}
log(x)
```

```{r}
sum(x)
```

```{r}
y <- 21:30
```

```{r}
x+y
```

```{r}
x^2+y^2   
```

```{r}
mean(x+y) 
```
Let's try something strange:

```{r}
c(1, 2, 3) + c(1, 2, 3, 4)
```
so R notices that we are trying to add a vector of length 3 to a vector of length 4. This should not work, but it actually does!

When it runs out of values in the first vector, R simply starts all over again. 

In general this is more likely a mistake by you, check that this is what you really wanted to do!

