---
header-includes: \usepackage{color}
                 \usepackage{float}
output:
  pdf_document:
    fig_caption: no
  html_document: default
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("R/setup.rmd.R", local=TRUE)
setup.rmd(local.env=environment())
```
`r hl()$basefontsize()`
`r hl()$style()`

## R Markdown, HTML and Latex

### R Markdown

R Markdown is a program for making dynamic documents with R. An R Markdown document is written in *markdown*, an easy-to-write plain text format with the file extension .Rmd. It can contain chunks of embedded R code. It has a number of great features:

-  easy syntax for a number of basic objects  
-  code and output are in the same place and so are always synced  
-  several output formats (html, latex, word)

In recent years I (along with many others) who work a lot with R have made Rmarkdown the basic way to work with R. So when I work on a new project I immediately start a corresponding R markdown document. 

`r hl()$hr()`

to start writing an R Markdown document open RStudio, File > New File > R Markdown. You can type in the title and some other things.

The default document starts like this:

\-\-\-  
title: "My first R Markdown Document"  
author: "Dr. Wolfgang Rolke"  
date: "April 1, 2018"  
output: html_document  
\-\-\-

This follows a syntax called YAML (also used by other programs). There are other things that can be put here as well, or you can erase all of it.

YAML stands for *Yet Another Markup Language*. It has become a standard for many computer languages to describe different configurations. For details go to [yaml.org](http://yaml.org)

Then there is other stuff you should erase. Next File > Save. Give the document a name with the extension .Rmd

### Basic R Markdown Syntax

for a list of the basic syntax go to

[https://rmarkdown.rstudio.com/articles_intro.html](https://rmarkdown.rstudio.com/articles_intro.html) 

or to 

[https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet)


#### Embedded Code

There are two ways to include code chunks (yes, that's what they are called!) into an R Markdown document:

####  stand alone code

simultaneously enter CTRL-ALT-i and you will see this:


\```{r}

\```



you can now enter any R code you like:

\```{r}  
x <- rnorm(10)  
mean(x)  
\```

which will appear in the final document as

```{r eval=FALSE}
x <- rnorm(10)
mean(x)
```

Actually, it will be like this:

```{r}
x<-rnorm(10)
mean(x)
```

so we can see the result of the R calculation as well. The reason it didn't appear like this before was that I added the argument eval=FALSE:

\```{r eval=FALSE}  
x<-rnorm(10)  
mean(x)  
\```

which keeps the code chunk from actually executing (aka *eval*uating). This is useful if the code takes along time to run, or if you want to show code that is actually faulty, or ...

there are a number of useful arguments:

-  eval=FALSE  (shows but doesn't run the code)  
-  eval=2:5 (shows all the code but only runs lines 2 to 5)
-  echo=FALSE  (the code chunk is run but does not appear in the document)  
-  echo=2:5 (shows only code on lines 2 to 5)  
-  warning=FALSE  (warnings are not shown)  
-  message=FALSE (messages are not shown)  
-  cache=TRUE  (code is run only if there has been a change, useful for lengthy calculations)  
-  error=TRUE  (if there are errors in the code R normaly terminates the parsing (executing) of the markdown document. With this argument it will ignore the error, which helps with debugging)  
-  engine='Rcpp' (to include C++ code)

####  inline code 

here is a bit of text:

... and so the mean was `r mean(x)`.

Now I didn't type in the number, it was done with the chunk `` `r
mean(x)` ``.

`r hl()$hr()`

Many of these options can be set globally, so they are active for the whole document. This is useful so you don't have to type them in every time. I have the following code chunk at the beginning of all my Rmd:

```{r eval=FALSE}
library(knitr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
```

We have already seen the message and warning options. The other one puts any figure in the middle of the page and sizes it nicely.

If you have to override these defaults just include that in the specific chunk.

### Creating Output

To create the output you have to "knit" the document. This is done by clicking on the *knit* button above. If you click on the arrow you can change the output format.

### HTML vs Latex(Pdf)

In order to knit to pdf you have to install a latex interpreter. My suggestion is to use [Miktex](https://miktex.org/), but if you already have one installed it might work as well.

There are several advantages / disadvantages to each output format:

-  HTML is much faster  
-  HTML looks good on a webpage, pdf looks good on paper  
-  HTML needs an internet connection to display math, pdf does not  
-  HTML can use both html and latex syntax, pdf works only with latex (and a little bit of html)

I generally use HTML when writting a document, and use pdf only when everything else is done. There is one problem with this, namely that a document might well knit ok to HTML but give an error message when knitting to pdf. Moreover, those error messages are weird! Not even the line numbers are anywhere near right. So it's not a bad idea to also knit to pdf every now and then.

### Tables

One of the more complicated things to do in R Markdown is tables. For a nice illustration look at

[https://stackoverflow.com/questions/19997242/simple-manual-rmarkdown-tables-that-look-good-in-html-pdf-and-docx](https://stackoverflow.com/questions/19997242/simple-manual-rmarkdown-tables-that-look-good-in-html-pdf-and-docx)

My preference is to generate a data frame and the use the *kable* function:

```{r}
Gender <- c("Male", "Male", "Female")
Age <- c(20, 21, 19)
knitr::kable(data.frame(Gender, Age))
```

probably with the argument echo=FALSE so only the table is visible.

It is also possible to use HTML code to make a table:

```{r echo=FALSE}
txt <- c(" <table border=\"1\">\n",
"<tr><th>Gender</th><th>Age</th></tr>\n",
"<tr><td>Male</td><td>20</td></tr>\n",
"<tr><td>Male</td><td>21</td></tr>\n",
"<tr><td>Female</td><td>19</td></tr>\n",
"</table>")
cat(txt)
```

will look like this in HTML:

<table border="1">
<tr><th>Gender</th><th>Age</th></tr>
<tr><td>Male</td><td>20</td></tr>
<tr><td>Male</td><td>21</td></tr>
<tr><td>Female</td><td>19</td></tr>
</table>

but won't look like anything in pdf. 

The corresponding latex table will look good in pdf but not in HTML!

### LATEX

You have not worked with latex (read: latek) before? Here is your chance to learn. It is well worthwhile, latex is the standard document wordprocessor for science. And once you get used to it, it is WAY better and easier than (say) Word. 

Because latex code will generally display correctly in an html document but html will not in a latex document I suggest to stick as much as possible with latex.

A nice list of common symbols is found on [https://artofproblemsolving.com/wiki/index.php/LaTeX:Symbols](https://artofproblemsolving.com/wiki/index.php/LaTeX:Symbols).


#### Multiline math

say you want the following in your document:

$$
\begin{aligned}
&E[X] = \int_{-\infty}^{\infty} xf(x) dx    = \\
&\int_{0}^{1} x dx    = \frac12 x^2 |_0^1 = \frac12
\end{aligned}
$$

for this to display correctly in HTML and PDF you need to use the format

```{r, echo=FALSE}
txt <- c(" $$\n",
         "\\begin{aligned}\n",
         "&E[X] = \\int_{-\\infty}^{\\infty} xf(x) dx=\\\\ \n",
         "&\\int_{0}^{1} x dx    = \\frac12 x^2 |_0^1 = \\frac12 \n",
          "\\end{aligned}\n",
          "$$"
         )
cat(txt)
```

By default when you knit to pdf the intermediate latex file is deleted. If you want to keep it, maybe so you can change it in a latex editor, use the following in the YAML header:

output:  
&nbsp;&nbsp;pdf_document:  
&nbsp;&nbsp;&nbsp;&nbsp;keep_tex: true

notice the spaces before the text, they are needed!



