---
title: Nonparametric Regression, Smoothing
header-includes: \usepackage{color}
output:
  html_document: default
  pdf_document:
    fig_caption: no
---
<style>
table, th, td { text-align:right; }
th, td {padding: 10px;}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
whichcomp <- strsplit(getwd(),"/")[[1]][3]
load(paste0("c:/users/", whichcomp, "/Dropbox/teaching/Resma3/Resma3.RData"))
library(knitr)
library(wolfr)
opts_chunk$set(fig.width=6, fig.align = "center", 
      out.width = "70%", warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
```
`r hl()$basefontsize()`

Consider the following graph:

```{r}
ggplot(wine, 
       aes(Wine.Consumption, Heart.Disease.Deaths)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

Notice that this is the default of geom_smoooth, if you want the least squares line you need to use the argument method="lm".

What is this curve?

A completely different approach to fitting non-linear problems is via local and non-parametric regression. Say we have a model of the form $y=f(x)$ for some function $f$, and we want to estimate $y_0=f(x_0)$. In this type of fitting procedure $y_0$ is estimated using only x values close to $x_0$, or at least the contribution of x values close to $x_0$ is weighted more heavily than others.

One popular method for smoothing is the function *loess*. It works as follows:

1) Find the k nearest neighbors of $x_0$, which constitute a neighborhood $N(x_0)$. The number of neighbors k is specified as a percentage of the total number of points in the dataset. This percentage is called the span and is a tuning parameter of the method.

2) Calculate the largest distance $D(x_0)$ between $x_0$ and another point in the neighborhood.

3) Assign weights to each point in $N(x_0)$ using the tri-cube weight function:

$$
\begin{aligned}
&W(\frac{x_0-x_1}{\Delta(x_0)} )  
\end{aligned}
$$
where

$$
W(u)=(1-u^3)^3 I_{[0,1]}(u)
$$
4) Calculate the weighted least squares fit of $x_0$ on the neighborhood $N(x_0)$. 

As we saw above, when it is supposed to be added to a ggplot it is easily done with the geom_smooth() command. The span can be changed with


```{r}
ggplot(wine, 
       aes(Wine.Consumption, Heart.Disease.Deaths)) +
  geom_point() +
  geom_smooth(se=FALSE, span=0.3, color="green") +
  geom_smooth(se=FALSE) +
  geom_smooth(se=FALSE, span=1.3, color="red") 
```

so a larger value of span leads to a smoother curve. The default is 0.75.

To get actual predictions do this:

```{r}
fit <- loess(Heart.Disease.Deaths~Wine.Consumption,
          data=wine)
predict(fit, data.frame(Wine.Consumption=5))
```

There are a number of nonparametric regression methods implemented in R:

-  **ksmooth** finds the *Nadaraya-Watson* kernel regression estimate, which is of the form

$$
\hat{y}_i=\frac{\sum_{j=1}^{n}y_iK(\frac{x_i-x_j}{h})}{\sum_{j=1}^{n}K(\frac{x_i-x_j}{h})}
$$
 
here $K$ is the kernel function, usually one of the following:

-  Normal density: $K(x)=\frac1{\sqrt{2 \pi}}\exp (-\frac12 x^2)$

-  Tukey's Biweight: $K(x)=\frac{15}{16}(1-x^2)^2I_{[0,1]}(x)$

-  Epanechnikov: $K(x)=\frac{3}{4}(1-x^2)I_{[0,1]}(x)$ 

and $h$ is the equivalent to *span* in loess, a tuning parameter.

-  **smooth.spline** fits a *cubic smoothing spline*.

Splines are smooth piecewise polynomial functions often used in numerical analysis. Cubic splines specifically use polynomials up to degree 3. The cubic functions change at points called *knots* and are chosen such that the whole function is continuous.

Let's consider the following artificial example: we generate some data, using the model

$$
\begin{aligned}
&y=5-2x+0.35x^2-0.01x^3 \text{, } x<5 \\
&y=2.5 \text {, } x>5 
\end{aligned}
$$
```{r}
x <- sort(runif(100, 0, 10))
df <- data.frame(x=x,
     y=ifelse(x<5, 5-2*x+0.35*x^2-0.01*x^3, 2.5)+ 
     rnorm(100, 0, 0.25))
plt <- ggplot(df, aes(x, y)) + geom_point()
newx <- seq(0, 10, length=100)
df.true <- data.frame(x=newx,
     y=ifelse(newx<5, 
              5-2*newx+0.35*newx^2-0.01*newx^3, 2.5))
plt + geom_line(data=df.true,  aes(x, y))
```

we wish to fit cubic splines to the dataset. That is we fit a model of the form

$$
y=\alpha_{0j}+\alpha_{1j}x+\alpha_{2j}x^2+\alpha_{3j}x^3
$$
if $x_{j-1}<x<x_j$ (with $x_0=-\infty$ and $x_{k+1}=\infty$). Here the $x_j$ are the *knots*. Sometimes these are also estimated from the data, but for now we keep it simple and assume we know $k=1$ and $x_1=5$. 

We want to estimate the $\alpha_{ij}$ via least squares but with the condition that the resulting function be continuous, which results in the condition

$$
\alpha_{0j}+\alpha_{1j}x_j+\alpha_{2j}x_j^2+\alpha_{3j}x_j^3 = \\
\alpha_{0j+1}+\alpha_{1j+1}x_j+\alpha_{2j+1}x_j^2+\alpha_{3j+1}x_j^3
$$

In general such *conditional optimization* problems can be solved with methods such as *Lagrange Multipliers*, but we don't need to worry about that, the routine *smooth.spline* takes care of it for us:

```{r}
fit <- smooth.spline(df$x, df$y, spar = 0.8)
df1 <- data.frame(predict(fit, 
        data.frame(x = seq(0, 10, length = 100))))
colnames(df1)[2] <- "y"
df1 <- rbind(df1, df.true)
df1$Which <- rep(c("Spline", "True"), each=100)
plt + geom_line(data=df1, aes(x, y, color=Which),
                size=1.2)
```

There is one drawback of this solution: it does not allow us to specify the exact location of the knot. Here is how to do this: first we introduce a new variable: $z=x-5$ if $x>5$ and 0 otherwise. Next we use *lm* to fit the model

$$
y=\gamma_{0}+\gamma_{1}x+\gamma_{2}x^2+\gamma_{3}x^3 +\\
\gamma_{4}z+\gamma_{5}z^2+\gamma_{6}z^3
$$
it is easy to see how to recover the $\alpha's$ from this fit.

```{r error=T}
x <- df$x
z <- ifelse(x < 5, 0, x - 5)
x2 <- x^2
x3 <- x^3
z2 <- z^2
z3 <- z^3
fit <- lm(df$y ~ x + x2 + x3 + z + z2 + z3)
g <- coef(fit)
a <- g[1:4]
b <- c(g[1] - g[5] * 5 + g[6] * 5^2 - g[7] * 5^3, g[2] + 
            g[5] - 2 * g[6] * 5 + 3 * g[7] * 5^2, g[3] +
         g[6] - 3 * g[7] * 5, g[4] + g[7])
print(rbind(a, b))
y <- rep(0, 100)
y[x <= 5] <- a[1] + a[2] * x[x <= 5] +
  a[3] * x[x <= 5]^2 + a[4] * x[x <= 5]^3
y[x > 5] <- b[1] + b[2] * x[x > 5] + b[3] * 
          x[x > 5]^2 + b[4] * x[x > 5]^3
df.tmp <- data.frame(x=x, y=y, Which=rep("lm", 100))
df1 <- rbind(df1, df.tmp)
plt + geom_line(data=df1, aes(x, y, color=Which),
            size=1.1)
```

Finally, let's add the loess and the ksmooth solutions as well:

```{r}
x <- seq(0, 10, length = 100)
fit <- loess(y~x, data=df)
df.loess <- data.frame(x=x, 
                 y=predict(fit, data.frame(x = x)),
                 Which=rep("loess", 100))
df.ksmooth <- data.frame(x=x, 
                 y=ksmooth(df$x, df$y, bandwidth = 2)$y,
                 Which=rep("ksmooth", 100))
df1 <- rbind(df1, df.loess, df.ksmooth)
plt + geom_line(data=df1, aes(x, y, color=Which))
```

Notice the different ways theses methods are called, mostly due to history of how and by whom they were added to R.

In general the choice of method is less important than the choice of:

## Bandwidth (Smoothing Parameter)

One of the most active research areas in Statistics in the last 20 years has been the search for a method to find the "optimal" bandwidth for a smooother. There are now a great number of methods to do this, unfortunately non of them is fully satisfactory. We will briefly look at one method which is one of the main contenders: *cross-validation*.

In order to find an "optimal" solution we need to first decide what "optimal" means. Here we will consider the MISE (mean integrated square error):

$$
\text{MISE }=E \left[]
\int ||\hat{f}(x;b)-f(x;b)||^2
\right]dx
$$
Of course ths MISE depends on the unknown function f, and so we need to estimate it from the data. This is done via cross-validation, that is using the observations $(x_1, y_1)$, .., $(x_{i-1},y_{i-1})$, $(x_{i+1},y_{i+1})$, .., $(x_n,y_n)$ to fit the curve at $x_i$ and get an estimate of $y_i$. Then you do this for all i and average over the error.

This specific description is often called *leave-one-out cross-validation*, for obvious reasons. One problem of this method is that for large datasets it is very computationally demanding.

cross-validation is already implemented in R for one of the smoothers, namely smooth.spline. If we do not specify a bandwith (spar or df) the routine will invoke the cross-validation procedure and choose the bandwidth automatically. 

```{r}
fit <- smooth.spline(df$x, df$y)
df2 <- data.frame(predict(fit, 
        data.frame(x = seq(0, 10, length = 100))))
colnames(df2)[2] <- "y"
plt + geom_line(data=df2, aes(x, y), 
                color="blue", size=1.2)
```

#### **Example**: Lunatics

Let's implement leave-one-out cross-validation for the loess method and apply it to the lunatics data:

```{r error=TRUE}
cr <- function(df, span) {
  n <- dim(df)[1]
  x <- df[[1]]
  y <- df[[2]]
  eps <- rep(0, n)
  for(i in 1:n) {
    fit <- loess(y[-i]~x[-i], span=span, surface="direct")
    yhat <- predict(fit, x[i])
    eps[i] <- (y[i]-yhat)^2
  }
  mean(eps)
}
```

```{r cache=TRUE}
span <- seq(0.6, 2, length=25)
eps <- span
for(i in 1:25) {
  eps[i] <- cr(lunatics[, 3:4], span[i])
}

```

```{r}
ggplot(data=data.frame(span=span, epsilon=eps),
       aes(span, epsilon)) +
  geom_line()
```

```{r}
span.cr <- span[eps==min(eps)]
span.cr
fit <- loess(Percent.at.Home~Distance, span=span.cr,
             data = lunatics[-13, ],
             surface="direct")
fit1 <- loess(Percent.at.Home~Distance, span=0.75,
             data = lunatics[-13, ],
             surface="direct")
x <- 0:100
y <- c(predict(fit, x), predict(fit1, x))
df <- data.frame(x=c(x, x), y,
          span=rep(c("cr", "default"), each=101))
ggplot(data=lunatics,
            aes(Distance, Percent.at.Home)) +
  geom_point() +
  geom_line(data=df,  aes(x, y, color=span),
            inherit.aes = FALSE)
```

and we see that the smoother curve looks better.

### Interval Estimation

How do we do interval estimation when using loess? As with the predict method for lm we can again get an estimate of the standard error by including *se=TRUE*. However, the predict command for loess does not have an  *interval"* argument. So how do we know whether these are confidence or prediction intervals?

Let's find out. For this we do a little simulation study. We generate 100 x uniform on [0,10] and then 100 y=10+3x+N(0,3). We fit the loess model (using span=0.75) and predict yhat at x=5.0, then we do the same using lm.

```{r cache=TRUE}
B <- 1000
se.loess = rep(0, B)
se.lm = rep(0, B)
for (i in 1:B) {
  x <- runif(100, 0, 10)
  y <- 10 + 3 * x + rnorm(100, 0, 3)
  se.loess[i] <-  predict(loess(y ~ x), 
                      newdata = data.frame(x = 5), 
                      se = TRUE)$se.fit
  se.lm[i] <- predict(lm(y ~ x), 
                      newdata = data.frame(x = 5), 
                      se = TRUE)$se.fit
}
cat("loess  error: ", sd(se.loess), "\n")
cat("lm  error: ", sd(se.lm), "\n")
```


We see that the errors in loess are larger (with a larger standard deviation) than those of lm. This to be expected, after all we are using a lot more information in lm, namely the exact model for the data and the exact distribution of the residuals. 

Notice also that 

```{r}
out <- predict(lm(y ~ x), newdata = data.frame(x = 5), 
       se = TRUE, interval="confidence")
round(out$fit[2:3], 1)
round(out$fit[1]+c(-1, 1)*qnorm(0.975)*out$se.fit, 1)
```

so we see that these errors give us confidence intervals. But what if we want prediction intervals?

We have the following equations for the errors:

$$
\begin{aligned}
&se_{fit}    = \hat{\sigma}\sqrt{\frac1n + \frac{(x-\bar{x})^2}{s_{xx}} } \\
&se_{pred}    = \hat{\sigma}\sqrt{1+\frac1n + \frac{(x-\bar{x})^2}{s_{xx}} } \\
\end{aligned}
$$
where $s_{xx}$ is the sum of squares and $\hat{\sigma}$ is an estimate of the standard deviation. It is part of the fit object with *as.numeric(fit[5])*. So

$$
\begin{aligned}
& \frac1n + \frac{(x-\bar{x})^2}{s_{xx}}   = 
\frac{se_{fit}}{\hat{\sigma}^2} \\
&se_{pred}    = \hat{\sigma}\sqrt{1+
\frac{se_{fit}^2}{\hat{\sigma}} } = \sqrt{\hat{\sigma}^2+se_{fit}^2 }\\
\end{aligned}
$$

and so we can find prediction intervals with

```{r}
fit <- loess(y ~ x)
yhat <- predict(fit)
sighat <- as.numeric(fit[5])
out <- predict(fit, 
                newdata = data.frame(x = 5), 
                se = TRUE)
se.pred <- sqrt(sighat^2 + out$se.fit^2)
round(out$fit[1]+c(-1, 1)*qnorm(0.975)*se.pred, 1)
```

#### **Example**: Lunatics

Find a $90\%$ prediction interval for the Percent.at.Home if the distance is 25miles, using the optimal span.

```{r}
fit <- loess(Percent.at.Home~Distance, span=span.cr,
             data = lunatics[-13, ],
             surface="direct")
sighat <- as.numeric(fit[5])
out <- predict(fit, 
                newdata = data.frame(Distance = 25), 
                se = TRUE)
se.pred <- sqrt(sighat^2 + out$se.fit^2)
round(out$fit[1]+c(-1, 1)*qnorm(0.95)*se.pred, 1)
```

### Confidence Bands

Let's have another look at geom_smooth:

```{r}
ggplot(data=lunatics[-13, ],
            aes(Distance, Percent.at.Home)) +
  geom_point() +
  geom_smooth(span=span.cr)
```

what is this gray band? It is a shaded area between the lower and the upper $95\%$ confidence intervals. We can recreate it ourselves:

```{r}
fit <- loess(Percent.at.Home~Distance, span=span.cr,
             data = lunatics[-13, ],
             surface="direct")
x <- 1:99
out <- predict(fit, 
                newdata = data.frame(Distance = x), 
                se = TRUE)
low <- out$fit-qnorm(0.975)*out$se.fit
high <- out$fit+qnorm(0.975)*out$se.fit
df.low <- data.frame(x=x, y=low)
df.high <- data.frame(x=x, y=high)
ggplot(data=lunatics[-13,],
            aes(Distance, Percent.at.Home)) +
  geom_point() +
  geom_smooth(method="loess", span=span.cr) +
  geom_line(data=df.low, aes(x, y), color="red")+
  geom_line(data=df.high, aes(x, y), color="red")
```

almost, except that geom_smooth does some additional adjustments. We won't worry about that for now.

How about if we want the bands to show prediction intervals?

```{r}
sighat <- as.numeric(fit[5])
se.pred <- sqrt(sighat^2 + out$se.fit^2)
low <- out$fit-qnorm(0.975)*se.pred
low[low<0] <- 0
high <- out$fit+qnorm(0.975)*se.pred
high[high>100] <- 100
df.low <- data.frame(x=x, y=low)
df.high <- data.frame(x=x, y=high)
ggplot(data=lunatics[-13,],
            aes(Distance, Percent.at.Home)) +
  geom_point() +
  geom_line(data=df.low, aes(x, y), color="red")+
  geom_line(data=df.high, aes(x, y), color="red")
```

or a bit nicer:

```{r}
df1 <- data.frame(x, ymin=low, ymax=high)
ggplot(data=lunatics[-13,],
            aes(Distance, Percent.at.Home)) +
  geom_point() +
  geom_ribbon(data=df1, 
              aes(x=x, ymin=ymin, ymax=ymax),
              alpha=0.2,
              inherit.aes = FALSE)
```

There is however an issue with these bands. From our recreation it is clear that they are *pointwise* confidence intervals, that is each is a $95\%$ confidence interval for each x value. However, psychologically most people will look at them and interpret them as *simultaneous* confidence bands. That is, the $95\%$ applies to the collection of intervals, not each interval alone. 

Say we have $n$ data points, and we find a $(1-\alpha)100\%$ confidence interval at each. If they are all independent we find

$$
\begin{aligned}
&P( \text{at least one interval is wrong})    = \\
&1-P( \text{no interval is wrong})    = \\
&1-P\left( \cap_{k=1}^n I_k \text{ is right} \right)    = \\
&1-\prod_{k=1}^nP\left(  I_k \text{ is right} \right)    = \\
&1-\prod_{k=1}^n(1-\alpha)=1-(1-\alpha)^n
\end{aligned}
$$
and this goes to 1 as n grows larger. 

To make matters worse, in a regression case intervals at neighboring points are clearly not independent, so we don't even know what the true simultaneous coverage might be. 

Personally I am very reluctant to add such bands to graphs, but they are quite popular in many fields.